\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-17 jasa/paper-jasa.tex
introduce Omega_u notation

% As a concrete example, consider searches for new particles at the Large Hadron
% Collider. The simulator that is sampling from $p(x|\theta)$ is based on quantum
% field theory, a detailed simulation of the particle detector, and data
% processing algorithms that transform raw sensor data into the feature vector
% $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}.
%
% The ATLAS and CMS experiments have published  hundreds papers where the final
% result was formulated as a hypothesis test or confidence interval using a
% generalized  likelihood ratio test~\citep{Cowan:2010js}. This includes the
% discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and
% subsequent measurement of its properties.
%
% The bulk of the likelihood ratio tests at the LHC are based on the distribution
% of a single event-level feature that discriminates between a hypothesized
% process of interest (labeled \textit{signal}) and various other processes
% (labeled \textit{background}). Typically,  pseudo-data from the simulator are
% used to approximate the density at various parameter points, and an
% interpolation algorithm is used to approximate the parametrized
% model~\citep{Cranmer:2012sba}.
%
% To improve the statistical power of these tests, hundreds of these searches have
% utilized supervised learning to train discriminative classifiers that take
% commonly used. Recently, there has been progress in using deep
% networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons
% learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.

As a concrete example, let us consider searches for new particles at the Large
Hadron Collider (LHC). The simulator that is sampling from $p(\mathbf{x}|\theta)$ is
based on quantum field theory, a detailed simulation of the particle detector,
and data processing algorithms that transform raw sensor data into the feature
vector $\mathbf{x}$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
experiments have published  hundreds of papers where the final result was
formulated as a hypothesis test or confidence interval using a generalized
likelihood ratio test~\citep{Cowan:2010js}, including most notably the discovery
of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and subsequent
measurement of its properties. The bulk of the likelihood ratio tests at the LHC
are based on the distribution of a single event-level feature that discriminates
between a hypothesized process of interest (labeled \textit{signal}) and various
other processes (labeled \textit{background}).
Typically, data generated from the
simulator are used to approximate the density at various parameter points, and
an interpolation algorithm is used to approximate the parameterized
model~\citep{Cranmer:2012sba}. In order to improve the statistical power
of these tests, hundreds of these searches have already been using supervised
learning to train classifiers to discriminate between two two discrete hypotheses based
on a high dimensional feature vector $\mathbf{x}$. The results of this paper outline
how to extend the use of discriminative classifiers for composite hypotheses (parametrized by $\theta$)
in a way
that fits naturally into the established likelihood based inference techniques.

% While classification accuracy can lead to optimal approaches for simple
% hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of
% parameter estimation or composite hypothesis tests with nuisance parameters. As
% noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they
% assume that the selector with the highest classification accuracy will yield a
% mass measurement with the smallest statistical uncertainty.'' The key
% distinction is that evaluating the loss for classification is composed of many
% per-event operations, while evaluating the loss for a mass measurement (e.g. the
% variance of an estimator for the mass parameter) is a per-experiment operation
% involving a data set with many events. They went on to demonstrate a
% computationally intensive stochastic optimization technique based on the
% per-experiment loss out performed the two stage selection-estimation process.
%
% The initial motivation for this work was to extend the typical usage of
% discriminative classifiers in HEP to be robust to nuisance parameters in the
% simulators. The scope of the result expanded once it became clear that this
% offers a way to approximate the likelihood function $p(x|\theta)$ in what is
% typically considered the likelihood-free setting. This approach is complementary
% to ABC as it does not require a prior over the parameters and can also be used
% in the classical (frequentist) setting. A strength of this approach is that it
% separates the quality of the approximation of the target likelihood from the
% quality of the calibration.  In Section~\ref{S:Related} we discuss the scheme
% sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a
% dimensionality reduction map to aid in the estimation of the likelihood
% function.

The rest of the paper is organized as follows. In
Section~\ref{sec:likelihood-ratio-tests}, we first introduce the likelihood
ratio test statistic in the setting of simple hypothesis testing, and then
outline how it can be computed exactly using calibrated classifiers.
In Section~\ref{sec:generalized-likelihood-ratio}, we generalize the proposed
approach to the case of composite hypothesis testing and discuss directions for
approximating the statistic efficiently. We then illustrate the proposed
method in Section~\ref{sec:examples} and outline how it could improve
statistical analysis within the field of high energy physics. Related work
and conclusions are finally presented in Sections~\ref{sec:related} and \ref{sec:conclusions}.

@@

% \subsection{Notation and Assumptions}
%
% We use the following notation:
% \begin{itemize}
%  \item $x$: a vector of features for an event
%  \item $D$: a data set of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
%  \item $\theta$: parameters of a statistical model
% \item $p(x| \theta)$:  probability density  (simulation-based model) for $x$ given $\theta$
% %\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
% \item $y$: a class label used for training a classifier.
% \item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
% %\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
% \item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$
% \end{itemize}
% We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.

@@

% The main result of this paper is to generalize the observation that one can form an equivalent test based on
% %\begin{equation}
% %T'(D) = \prod_{e=1}^n \frac{ p(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ p(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
% %\end{equation}
% \begin{equation}\label{eq:equivLRtest}
% T'(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
% \end{equation}
% if
% \begin{equation}\label{eq:montonic}
% s_e = s(x_e; \theta_0, \theta_1) = m\left(\, p(x_e|\theta_0) / p(x_e|\theta_1) \,\right) \;
% %s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)} \right) \;
% \end{equation}
% where $m$ is any strictly increasing or decreasing function. This result will be proven below.
% This allows us to recast the original likelihood ratio test into an alternate form in which supervised learning is used to train the discriminative classifier $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,u=0)$ generated
% from $p(x|\theta_0)$ and $(x,u=1)$ generated from $p(x|\theta_1)$. In Section~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the classifier  parametrized in terms of $(\theta_0, \theta_1)$.
%
% Here we see that the original goal for simple hypothesis testing (i.e. to make a decision to accept or reject the null hypothesis based on the entire data set $D$) has been reformulated into a per-event classification problem. This follows from the fact that we assume the $x_e$ to be i.i.d.

The main result of this paper is to
generalize the observation that one can form a test statistic
\begin{equation}\label{eqn:likelihood-ratio-test-equiv}
\lambda'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{U}(u=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{U}(u=s(\mathbf{x}) | \theta_1)}
\end{equation}
that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
of variable $\mathbf{U} = s(\mathbf{X})$ is based
on a (parameterized) function $s$ that is strictly monotonic with the density ratio
\begin{equation}
r(\mathbf{x};\theta_0, \theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
\end{equation}
As derived below, this allows to recast the original likelihood ratio test into
an alternate form in which supervised learning can be used to build
$s(\mathbf{x})$ as a discriminative classifier.  In
Section~\ref{sec:generalized-likelihood-ratio} we extend this result to
generalized likelihood ratio tests, where it will be useful to have the
classifier decision function $s$ parameterized in terms of $(\theta_0, \theta_1)$.

@@

\begin{proof}
Starting from the definition of the probability density function, we have
\begin{align}
% p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
% &= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
% &= \int_{\mathbb{R}^p} \frac{d}{dy} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
&= \int_{\mathbb{R}^p} \delta(u - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime
\end{align}
Intuitively, this expression can be understood as the integral
over all $\mathbf{x}^\prime \in \mathbb{R}^p$ such that $s(\mathbf{x}^\prime) = u$, as picked
by the Dirac $\delta$ function. Given Theorem 6.1.5 of \citet{Hrmander1990},
it further comes
\begin{equation}
p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) = \int_{\mathbf{x}^\prime \in \Omega_u } \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
\end{equation}
where $\Omega_u = \{x^\prime : s(\mathbf{x}^\prime)=u\}$, $|\nabla s(\mathbf{x}^\prime)| = \sqrt{\sum_{i=1}^p |\frac{\partial}{\partial x_i} s(\mathbf{x}^\prime)|^2}$
and where $dS_{\mathbf{x}^\prime}$ is the Euclidean surface measure on $\Omega_u$.
Also, since $s(\mathbf{x})$ is monotonic with
$r(\mathbf{X}; \theta_0, \theta_1)$,
there exists an invertible function $m:\mathbb{R}^+ \mapsto \mathbb{R}$ such
that $s(\mathbf{x}) = m(r(\mathbf{X}; \theta_0, \theta_1))$.
In particular, we have
\begin{align}
\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} &= m^{-1}(s(\mathbf{x})) \nonumber \\
p_\mathbf{X}(\mathbf{x}|\theta_0) &= m^{-1}(s(\mathbf{x})) p_\mathbf{X}(\mathbf{x}|\theta_1) \label{eqn:mapping}
\end{align}
Combining equations \ref{eqn:hormander} and \ref{eqn:mapping}, the density ratio $r(\mathbf{X}; \theta_0, \theta_1)$ can be pulled out of the integral, resulting in
\begin{align}
p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
&= \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(u) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
&= m^{-1}(s(\mathbf{x})) \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}. \label{eqn:factorization}
\end{align}
Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)$, finally yielding
\begin{align}
\frac{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0)}{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}. \label{eqn:jacob}
\end{align}
\end{proof}

@@

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the decision
function
\begin{equation}\label{eqn:best-s-clf}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies the conditions of
Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
In other words, supervised learning yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio,  not only for $m(r(\mathbf{x})) = (1 +
r(\mathbf{x}))^{-1}$.
Equivalently,
Theorem~\ref{thm:ratio-equivalence} shows that in the case that we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_{\mathbf{U}}(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
For this reason, the proposed method is expected to be more robust than
directly using ${(1 - s(\mathbf{x}))}/{s(\mathbf{x})}$ as an approximate of
$r(\mathbf{x})$ (which indeed converges towards $r(x)$ when $s(x)$ tends to $s^*(x)$).

@@

An alternative approach for calibration is to approximate the density ratio
$r(\hat{s}(\mathbf{x}))$ directly.
For example, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
of Eqn.~\ref{eqn:best-s-clf}, can be used for calibration. This is done by inverting the relationship ${r}(\mathbf{x}) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\hat{s}(\mathbf{x})) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
Additionally, Sec.~\ref{sec:related} describes related work in which the density ratio $\hat{r}(\mathbf{x})$ is estimated directly on the feature space $\mathcal{X}$.

@@

% - typical usage in HEP with a concrete experimental example.
%   * illustrate the method for several classifiers (new)
%    * illustrate the method for several calibration routines (new)
%    * compare with other methods (new)
% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parametrized classifiers)
%   discuss about focusing the capacity of the classifier
% - measure of particles properties, with a concrete example. Comparison with other likelihood-free inference methods (new)

>>>>

% As a concrete example, consider searches for new particles at the Large Hadron
% Collider. The simulator that is sampling from $p(x|\theta)$ is based on quantum
% field theory, a detailed simulation of the particle detector, and data
% processing algorithms that transform raw sensor data into the feature vector
% $x$~\citep{Sjostrand:2006za,Agostinelli:2002hh}.
%
% The ATLAS and CMS experiments have published  hundreds papers where the final
% result was formulated as a hypothesis test or confidence interval using a
% generalized  likelihood ratio test~\citep{Cowan:2010js}. This includes the
% discovery of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and
% subsequent measurement of its properties.
%
% The bulk of the likelihood ratio tests at the LHC are based on the distribution
% of a single event-level feature that discriminates between a hypothesized
% process of interest (labeled \textit{signal}) and various other processes
% (labeled \textit{background}). Typically,  pseudo-data from the simulator are
% used to approximate the density at various parameter points, and an
% interpolation algorithm is used to approximate the parameterized
% model~\citep{Cranmer:2012sba}.
%
% To improve the statistical power of these tests, hundreds of these searches have
% utilized supervised learning to train discriminative classifiers that take
% commonly used. Recently, there has been progress in using deep
% networks~\citep{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons
% learned during HiggsML~\citep{HepML}, the largest Kaggle challenge in history.

As a concrete example, let us consider searches for new particles at the Large
Hadron Collider (LHC). The simulator that is sampling from $p(\mathbf{x}|\theta)$ is
based on quantum field theory, a detailed simulation of the particle detector,
and data processing algorithms that transform raw sensor data into the feature
vector $\mathbf{x}$~\citep{Sjostrand:2006za,Agostinelli:2002hh}. The ATLAS and CMS
experiments have published  hundreds of papers where the final result was
formulated as a hypothesis test or confidence interval using a generalized
likelihood ratio test~\citep{Cowan:2010js}, including most notably the discovery
of the Higgs boson~\citep{Aad:2012tfa,Chatrchyan:2012ufa} and subsequent
measurement of its properties. The bulk of the likelihood ratio tests at the LHC
are based on the distribution of a single event-level feature that discriminates
between a hypothesized process of interest (labeled \textit{signal}) and various
other processes (labeled \textit{background}).
Typically, data generated from the
simulator are used to approximate the density at various parameter points, and
an interpolation algorithm is used to approximate the parameterized
model~\citep{Cranmer:2012sba}. In order to improve the statistical power
of these tests, hundreds of these searches have already been using supervised
learning to train classifiers to discriminate between two two discrete hypotheses based
on a high dimensional feature vector $\mathbf{x}$. The results of this paper outline
how to extend the use of discriminative classifiers for composite hypotheses (parameterized by $\theta$)
in a way
that fits naturally into the established likelihood based inference techniques.

% While classification accuracy can lead to optimal approaches for simple
% hypothesis tests~\citep{Dempster1965}, that is no longer true in the context of
% parameter estimation or composite hypothesis tests with nuisance parameters. As
% noted in ~\citep{Whiteson:2006ws}, ``such methods are suboptimal because they
% assume that the selector with the highest classification accuracy will yield a
% mass measurement with the smallest statistical uncertainty.'' The key
% distinction is that evaluating the loss for classification is composed of many
% per-event operations, while evaluating the loss for a mass measurement (e.g. the
% variance of an estimator for the mass parameter) is a per-experiment operation
% involving a data set with many events. They went on to demonstrate a
% computationally intensive stochastic optimization technique based on the
% per-experiment loss out performed the two stage selection-estimation process.
%
% The initial motivation for this work was to extend the typical usage of
% discriminative classifiers in HEP to be robust to nuisance parameters in the
% simulators. The scope of the result expanded once it became clear that this
% offers a way to approximate the likelihood function $p(x|\theta)$ in what is
% typically considered the likelihood-free setting. This approach is complementary
% to ABC as it does not require a prior over the parameters and can also be used
% in the classical (frequentist) setting. A strength of this approach is that it
% separates the quality of the approximation of the target likelihood from the
% quality of the calibration.  In Sec.~\ref{S:Related} we discuss the scheme
% sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a
% dimensionality reduction map to aid in the estimation of the likelihood
% function.

The rest of the paper is organized as follows. In
Sec.~\ref{sec:likelihood-ratio-tests}, we first introduce the likelihood
ratio test statistic in the setting of simple hypothesis testing, and then
outline how it can be computed exactly using calibrated classifiers.
In Sec.~\ref{sec:generalized-likelihood-ratio}, we generalize the proposed
approach to the case of composite hypothesis testing and discuss directions for
approximating the statistic efficiently. We then illustrate the proposed
method in Sec.~\ref{sec:examples} and outline how it could improve
statistical analysis within the field of high energy physics. Related work
and conclusions are finally presented in Sections~\ref{sec:related} and \ref{sec:conclusions}.

@@

% \subsection{Notation and Assumptions}
%
% We use the following notation:
% \begin{itemize}
%  \item $x$: a vector of features for an event
%  \item $D$: a data set of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
%  \item $\theta$: parameters of a statistical model
% \item $p(x| \theta)$:  probability density  (simulation-based model) for $x$ given $\theta$
% %\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
% \item $y$: a class label used for training a classifier.
% \item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parameterized by $\theta_0$ and $\theta_1$
% %\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
% \item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$
% \end{itemize}
% We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.

@@

% The main result of this paper is to generalize the observation that one can form an equivalent test based on
% %\begin{equation}
% %T'(D) = \prod_{e=1}^n \frac{ p(\,s(x_e; \theta_1, \theta_0) \mid \theta_1)}{ p(\,s(x_e; \theta_1, \theta_0)\mid\theta_0)}
% %\end{equation}
% \begin{equation}\label{eq:equivLRtest}
% T'(D; \theta_0, \theta_1) = \prod_{e=1}^n \frac{ p(s_e | \theta_0)}{ p(s_e | \theta_1)}
% \end{equation}
% if
% \begin{equation}\label{eq:montonic}
% s_e = s(x_e; \theta_0, \theta_1) = m\left(\, p(x_e|\theta_0) / p(x_e|\theta_1) \,\right) \;
% %s_e = s(x_e; \theta_0, \theta_1) = m\left(\frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)} \right) \;
% \end{equation}
% where $m$ is any strictly increasing or decreasing function. This result will be proven below.
% This allows us to recast the original likelihood ratio test into an alternate form in which supervised learning is used to train the discriminative classifier $s(x; \theta_0, \theta_1)$. The discriminative classifier can be trained with data $(x,u=0)$ generated
% from $p(x|\theta_0)$ and $(x,u=1)$ generated from $p(x|\theta_1)$. In Sec.~\ref{S:GLR} we extend this result to generalized likelihood ratio tests, where it will be useful to have the classifier  parameterized in terms of $(\theta_0, \theta_1)$.
%
% Here we see that the original goal for simple hypothesis testing (i.e. to make a decision to accept or reject the null hypothesis based on the entire data set $D$) has been reformulated into a per-event classification problem. This follows from the fact that we assume the $x_e$ to be i.i.d.

The main result of this paper is to
generalize the observation that one can form a test statistic
\begin{equation}\label{eqn:likelihood-ratio-test-equiv}
\lambda'({\cal D}; \theta_0, \theta_1) = \prod_{\mathbf{x} \in {\cal D}} \frac{ p_\mathbf{U}(u=s(\mathbf{x}) | \theta_0)}{ p_\mathbf{U}(u=s(\mathbf{x}) | \theta_1)}
\end{equation}
that is strictly equivalent to \ref{eqn:likelihood-ratio-test}, provided the change
of variable $\mathbf{U} = s(\mathbf{X})$ is based
on a (parameterized) function $s$ that is strictly monotonic with the density ratio
\begin{equation}
r(\mathbf{x};\theta_0, \theta_1) = \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.
\end{equation}
As derived below, this allows to recast the original likelihood ratio test into
an alternate form in which supervised learning can be used to build
$s(\mathbf{x})$ as a discriminative classifier.  In
Sec.~\ref{sec:generalized-likelihood-ratio} we extend this result to
generalized likelihood ratio tests, where it will be useful to have the
classifier decision function $s$ parameterized in terms of $(\theta_0, \theta_1)$.

@@

\begin{proof}
Starting from the definition of the probability density function, we have
\begin{align}
% p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
% &= \frac{d}{dy} \int_{\mathbb{R}^p} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
% &= \int_{\mathbb{R}^p} \frac{d}{dy} H(y - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\
&= \int_{\mathbb{R}^p} \delta(u - s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime
\end{align}
Intuitively, this expression can be understood as the integral
over all $\mathbf{x}^\prime \in \mathbb{R}^p$ such that $s(\mathbf{x}^\prime) = u$, as picked
by the Dirac $\delta$ function. Given Theorem 6.1.5 of \citet{Hrmander1990},
it further comes
\begin{equation}
p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) = \int_{\mathbf{x}^\prime \in \Omega_u } \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}
\end{equation}
where $\Omega_u = \{x^\prime : s(\mathbf{x}^\prime)=u\}$, $|\nabla s(\mathbf{x}^\prime)| = \sqrt{\sum_{i=1}^p |\frac{\partial}{\partial x_i} s(\mathbf{x}^\prime)|^2}$
and where $dS_{\mathbf{x}^\prime}$ is the Euclidean surface measure on $\Omega_u$.
Also, since $s(\mathbf{x})$ is monotonic with
$r(\mathbf{X}; \theta_0, \theta_1)$,
there exists an invertible function $m:\mathbb{R}^+ \mapsto \mathbb{R}$ such
that $s(\mathbf{x}) = m(r(\mathbf{X}; \theta_0, \theta_1))$.
In particular, we have
\begin{align}
\frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} &= m^{-1}(s(\mathbf{x})) \nonumber \\
p_\mathbf{X}(\mathbf{x}|\theta_0) &= m^{-1}(s(\mathbf{x})) p_\mathbf{X}(\mathbf{x}|\theta_1) \label{eqn:mapping}
\end{align}
Combining equations \ref{eqn:hormander} and \ref{eqn:mapping}, the density ratio $r(\mathbf{X}; \theta_0, \theta_1)$ can be pulled out of the integral, resulting in
\begin{align}
p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
&= \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(u) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
&= m^{-1}(s(\mathbf{x})) \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\
&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}. \label{eqn:factorization}
\end{align}
Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)$, finally yielding
\begin{align}
\frac{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0)}{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\Omega_u}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}. \label{eqn:jacob}
\end{align}
\end{proof}

@@

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the decision
function
\begin{equation}\label{eqn:best-s-clf}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies the conditions of
Thm.~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s}).
In other words, supervised learning yields a sufficient
procedure for Thm.~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Thm.~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio,  not only for $m(r(\mathbf{x})) = (1 +
r(\mathbf{x}))^{-1}$.
Equivalently,
Thm.~\ref{thm:ratio-equivalence} shows that in the case that we learn a
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_{\mathbf{U}}(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
For this reason, the proposed method is expected to be more robust than
directly using ${(1 - s(\mathbf{x}))}/{s(\mathbf{x})}$ as an approximate of
$r(\mathbf{x})$ (which indeed converges towards $r(x)$ when $s(x)$ tends to $s^*(x)$).

@@

An alternative approach for calibration is to approximate the density ratio
$r(\hat{s}(\mathbf{x}))$ directly.
For example, isotonic regression, which is commonly used to transform the classifier score  $\hat{s}(\mathbf{x})$ into $\hat{s}_\textrm{iso}(\mathbf{x})$ that more accurately reflect  the posterior probability $s^*(\mathbf{x})$
of Eqn.~\ref{eqn:best-s-clf}, can be used for calibration. This is done by inverting the relationship ${r}(\mathbf{x}) = (1-s^*(\mathbf{x}))/s^*(\mathbf{x})$ to obtain $\hat{r}(\hat{s}(\mathbf{x})) = (1-\hat{s}_\textrm{iso}(\mathbf{x}))/\hat{s}_\textrm{iso}(\mathbf{x})$.
Additionally, Sec.~\ref{sec:related} describes related work in which the ratio $\hat{r}(\mathbf{x})$ is estimated directly on the feature space $\mathcal{X}$.

@@

% - typical usage in HEP with a concrete experimental example.
%   * illustrate the method for several classifiers (new)
%    * illustrate the method for several calibration routines (new)
%    * compare with other methods (new)
% - ratio of mixtures and decomposition (special case of generalized likelihood rario tests when parameters only affect the mixture coefficients => no need for parameterized classifiers)
%   discuss about focusing the capacity of the classifier
% - measure of particles properties, with a concrete example. Comparison with other likelihood-free inference methods (new)
